Arrows for Parallel Computations

### 2017-05-19 paper/content/main.tex
generated pdf

\section{Introduction}
\label{sec:introduction}
\olcomment{todo, reuse 5.5, "Impact" at the end and more}

@@

HIT HERE REALLY STRONG

@@

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}}. For \ensuremath{\Varid{parEvalN}} this means that we can just apply the list of functions \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} to the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} by zipping them with the application operator \ensuremath{\mathbin{\$}}. % $
We then evaluate this lazy list \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} according to a \ensuremath{\Conid{Strategy}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} with the \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} operator. We construct this strategy with \ensuremath{\Varid{parList}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Conid{Strategy}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{\Varid{rdeepseq}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Conid{Strategy}\;\Varid{a}} where the latter is a strategy which evalutes to normal form. To ensure that programs that use \ensuremath{\Varid{parEvalN}} have the correct evaluation order, we annotate the computation with \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} which forces the compiler to not reorder multiple \ensuremath{\Varid{parEvalN}} computations. This is particularly necessary in circular communication topologies like in the \ensuremath{\Varid{torus}} or \ensuremath{\Varid{ring}} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \ensuremath{\Varid{pseq}} during testing for this paper.

%%%% BROKEN!
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\mathbf{let}\;\Varid{bs}\mathrel{=}\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as}{}\<[E]%
\\
\>[B]{}\hsindent{18}{}\<[18]%
\>[18]{}\mathbf{in}\;(\Varid{bs}\mathbin{`\Varid{using}`}\Varid{parList}\;\Varid{rdeepseq})\mathbin{`\Varid{pseq}`}\Varid{bs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

The most prominent instances of this interface are regular functions \ensuremath{(\to )} (Fig.~\ref{fig:arrowfn}),

@@

\caption{Arrow instance for regular functions}
\label{fig:arrowfn}
and the Kleisli type (Fig.~\ref{fig:arrowkleisli}).

@@

\caption{Definition of the Kleisli type and the corresponding arrow instance}
\label{fig:arrowkleisli}

@@

{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &

@@

{second\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\

@@

{(***)\label{fig:***Img}}{\includegraphics[width = 1.5in]{images/starstarstar}} &

@@

{(\&\&\&)\label{fig:&&&Img}}{\includegraphics[width = 1.5in]{images/andandand}}\

@@

	\caption{Syntactic sugar for arrows}

@@

With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \ensuremath{\Varid{first}}, called \ensuremath{\Varid{second}} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),

@@

\caption{The second combinator}
\label{fig:second}
the \ensuremath{\mathbin{*\!*\!*}} combinator which combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})

@@

\caption{The (***) combinator}
\label{fig:***}
and the \ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}} combinator that constructs an arrow which outputs two different values like \ensuremath{\mathbin{*\!*\!*}}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).

@@

\caption{The (\&\&\&) combinator}
\label{fig:&&&}
A short example given by Hughes on how to use this is \ensuremath{\Varid{add}} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.

@@

\caption{Add over arrows}
\label{fig:addArrows}

@@

Translating this into arrow terms gives us a new operator \ensuremath{\Varid{parEvalN}} that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to a parallel arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \ensuremath{\Varid{listApp}} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation).

@@

\caption{parEvalN Arrow combinator as a function}
\label{fig:parEvalNArrowFn}
With this definition of \ensuremath{\Varid{parEvalN}}, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow \ensuremath{\Varid{arr}} and we want this to be an interface for different backends, we introduce a new typeclass \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).

@@

\caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
\label{fig:parEvalNArrowTypeClass1}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs. normalform). For this reason we also introduce an additional \ensuremath{\Varid{conf}} parameter to the function. We also do not want \ensuremath{\Varid{conf}} to be a fixed type, as the configuration parameters can differ for different instances of \ensuremath{\Conid{ArrowParallel}}. So we add it to the type signature of the typeclass as well and get \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).

@@

\caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
\label{fig:parEvalNArrowTypeClassFinal}
Note that we don't require the \ensuremath{\Varid{conf}} parameter in every implementation. If it is not needed, we usually just default the \ensuremath{\Varid{conf}} type parameter to () and even blank it out in the parameter list of the implemented \ensuremath{\Varid{parEvalN}}, as we will see in the implementation of the Multicore and the ParMonad backend.

@@

The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \ensuremath{\Varid{withStrategy}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Varid{a}\to \Varid{a}} and \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} combinators from Multicore Haskell, where \ensuremath{\Varid{withStrategy}} is the same as \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} but with flipped parameters.

@@

\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()\;\mathbf{where}{}\<[E]%

@@

For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \ensuremath{\Conid{Strategy}} interface, we want to the user not to lose this ability because of using our API. We therefore introduce the \ensuremath{\Conid{Conf}\;\Varid{a}} data-type that simply wraps a \ensuremath{\Conid{Strategy}\;\Varid{a}} (Fig.~\ref{fig:confa}). We can't directly use the \ensuremath{\Conid{Strategy}\;\Varid{a}} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.

@@

\caption{Definition of Conf a}
\label{fig:confa}

@@

\subsubsection{Par Monad}
The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \ensuremath{\Varid{spawnP}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} function. The latter forks away the computation of a value and returns an \ensuremath{\Conid{IVar}} containing the result in the \ensuremath{\Conid{Par}} monad.

We therefore apply each function to its corresponding input value with \ensuremath{\Varid{listApp}} (Fig.~\ref{fig:listApp}) and then fork the computation away with \ensuremath{\Varid{arr}\;\Varid{spawnP}} inside a \ensuremath{\Varid{zipWithArr}} call. This yields a list \ensuremath{[\mskip1.5mu \Conid{Par}\;(\Conid{IVar}\;\Varid{b})\mskip1.5mu]}, which we then convert into \ensuremath{\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{sequenceA}}. In order to wait for the computation to finish, we map over the \ensuremath{\Conid{IVar}}s inside the ParMonad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})}. The result of this operation is a \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} from which we can finally remove the monad again by running \ensuremath{\Varid{arr}\;\Varid{runPar}} to get our output of \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
\\
\>[9]{}\hsindent{8}{}\<[17]%
\>[17]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{zipWithArr}\;(\Varid{app}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{spawnP})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{arr}\;\Varid{sequenceA}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{arr}\;\Varid{runPar}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$ %% formatting
\caption{ArrowParallel instance for the Par Monad backend}
\label{fig:ArrowParallelParMonad}
\end{figure}

>>>>

\section{Introduction}
\label{sec:introduction}
\olcomment{todo, reuse 5.5, and more}

@@

\olcomment{HIT HERE REALLY STRONG}

%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
We wrap parallel Haskells inside of our \ensuremath{\Conid{ArrowParallel}} interface, but why do we aim to abstract parallelism this way and what does this approach do better than the other parallel Haskells?
\begin{itemize}
	\item \textbf{Arrow API benefits}:
	With the \ensuremath{\Conid{ArrowParallel}} typeclass we do not lose any benefits of using arrows as \ensuremath{\Varid{parEvalN}} is just yet another arrow combinator. The resulting arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into our sequential programs without having to change anything.
	\item \textbf{Abstraction}:
	With the \ensuremath{\Conid{ArrowParallel}} typeclass, we abstracted all parallel implementation logic away from the business logic. This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell. So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual \ensuremath{\Conid{ArrowParallel}} instance.
\end{itemize}

@@

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}}. For \ensuremath{\Varid{parEvalN}} this means that we can just apply the list of functions \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} to the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} by zipping them with the application operator \ensuremath{\mathbin{\$}}. % $
We then evaluate this lazy list \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} according to a \ensuremath{\Conid{Strategy}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} with the \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} operator. We construct this strategy with \ensuremath{\Varid{parList}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Conid{Strategy}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{\Varid{rdeepseq}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Conid{Strategy}\;\Varid{a}} where the latter is a strategy which evalutes to normal form. To ensure that programs that use \ensuremath{\Varid{parEvalN}} have the correct evaluation order, we annotate the computation with \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} which forces the compiler to not reorder multiple \ensuremath{\Varid{parEvalN}} computations. This is particularly necessary in circular communication topologies like in the \ensuremath{\Varid{torus}} or \ensuremath{\Varid{ring}} skeleton that we will see in Section~\ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \ensuremath{\Varid{pseq}} during testing for this paper.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\mathbf{let}\;\Varid{bs}\mathrel{=}\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as}{}\<[E]%
\\
\>[B]{}\hsindent{18}{}\<[18]%
\>[18]{}\mathbf{in}\;(\Varid{bs}\mathbin{`\Varid{using}`}\Varid{parList}\;\Varid{rdeepseq})\mathbin{`\Varid{pseq}`}\Varid{bs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

@@

The most prominent instances of this interface are regular functions \ensuremath{(\to )}: %(Fig.~\ref{fig:arrowfn}),
%\begin{figure}[h]

@@

% \caption{Arrow instance for regular functions}
% \label{fig:arrowfn}
% \end{figure}
%%% on the verge of acceptable code
and the Kleisli type: %(Fig.~\ref{fig:arrowkleisli}).
% \begin{figure}[h]

@@

% \caption{Definition of the Kleisli type and the corresponding arrow instance}
% \label{fig:arrowkleisli}
% \end{figure}

@@

{\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &

@@

{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
\ensuremath{\Varid{first}} & \ensuremath{\Varid{second}} \\
\midrule

@@

{}{\includegraphics[width = 1.5in]{images/starstarstar}} &

@@

{}{\includegraphics[width = 1.5in]{images/andandand}}\\
\ensuremath{(\mathbin{*\!*\!*})}\label{fig:***Img} & \ensuremath{(\mathbin{\&\!\&\!\&})} \label{fig:&&&Img} \\

@@

	\caption{Visual depiction of syntactic sugar for Arrows.}

@@

With this typeclass in place, Hughes also defined some syntactic sugar: the functions \ensuremath{\Varid{second}}, \ensuremath{\mathbin{*\!*\!*}} and \ensuremath{\mathbin{\&\!\&\!\&}}. 
The mirrored version of \ensuremath{\Varid{first}}, called \ensuremath{\Varid{second}} (Fig.~\ref{fig:secondImg}% ,~\ref{fig:second}
) is:
% \begin{figure}[h]

@@

% \caption{The second combinator}
% \label{fig:second}
% \end{figure}
the \ensuremath{\mathbin{*\!*\!*}} combinator that combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, (Fig.\ref{fig:***Img}%
% ,~\ref{fig:***}
) is defined as
% \begin{figure}[h]

@@

% \caption{The (***) combinator}
% \label{fig:***}
% \end{figure}
and the \ensuremath{\mathbin{\&\!\&\!\&}} combinator that constructs an arrow which outputs two different values like \ensuremath{\mathbin{*\!*\!*}}, but takes only one input (Fig.~\ref{fig:&&&Img}% ,~\ref{fig:&&&}
) is:
% \begin{figure}[h]

@@

% \caption{The (\&\&\&) combinator}
% \label{fig:&&&}
% \end{figure}
A short example given by Hughes on how to use this is 
addition with arrows:
%|add| over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
% \begin{figure}[h]

@@

% \caption{Add over arrows}
% \label{fig:addArrows}
% \end{figure}

@@

Translating this into arrow terms gives us a new operator \ensuremath{\Varid{parEvalN}} that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to a parallel arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}. % (Fig.~\ref{fig:parEvalNArrowFn}) (
This combinator is similar to our utility function \ensuremath{\Varid{listApp}} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
% \begin{figure}[h]

@@

% \caption{parEvalN Arrow combinator as a function}
% \label{fig:parEvalNArrowFn}
% \end{figure}
With this definition of \ensuremath{\Varid{parEvalN}}, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow \ensuremath{\Varid{arr}} and we want this to be an interface for different backends, we introduce a new typeclass \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}} to host this combinator: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
% \begin{figure}[h]

@@

% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClass1}
% \end{figure}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs. normalform). For this reason we also introduce an additional \ensuremath{\Varid{conf}} parameter to the function. We also do not want \ensuremath{\Varid{conf}} to be a fixed type, as the configuration parameters can differ for different instances of \ensuremath{\Conid{ArrowParallel}}. So we add it to the type signature of the typeclass as well and get \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}}: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
% \begin{figure}[h]
\olcomment{\ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}} or \ensuremath{\Conid{ArrowParallel}\;\Varid{conf}\;\Varid{arr}\;\Varid{a}\;\Varid{b}}?}

@@

% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClassFinal}
% \end{figure}
Note that we don't require the \ensuremath{\Varid{conf}} parameter in every implementation. If it is not needed, we usually just default the \ensuremath{\Varid{conf}} type parameter to \ensuremath{()} and even blank it out in the parameter list of the implemented \ensuremath{\Varid{parEvalN}}, as we will see in the implementation of the Multicore and the \ensuremath{\Conid{Par}} Monad backend.

@@

The Multicore Haskell implementation of this class is implemented in a straightforward manner by using \ensuremath{\Varid{listApp}} from Appendix~\ref{utilfns} combined with the \ensuremath{\Varid{withStrategy}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Varid{a}\to \Varid{a}} and \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} combinators from Multicore Haskell, where \ensuremath{\Varid{withStrategy}} is the same as \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} but with flipped parameters.

@@

\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()\;\mathbf{where}{}\<[E]%

@@

For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \ensuremath{\Conid{Strategy}} interface, we want to the user not to lose this ability because of using our API. We therefore introduce the \ensuremath{\Conid{Conf}\;\Varid{a}} data-type that simply wraps a \ensuremath{\Conid{Strategy}\;\Varid{a}} (Fig.~\ref{fig:confa}). We can't directly use the \ensuremath{\Conid{Strategy}\;\Varid{a}} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances:
% \begin{figure}[h]

@@

% \caption{Definition of Conf a}
% \label{fig:confa}
% \end{figure}

@@

\subsubsection{\ensuremath{\Conid{Par}} Monad}
\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting}
The \ensuremath{\Conid{Par}} monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and \ensuremath{\Conid{Par}} monad's \ensuremath{\Varid{spawnP}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} function. The latter forks away the computation of a value and returns an \ensuremath{\Conid{IVar}} containing the result in the \ensuremath{\Conid{Par}} monad.

We therefore apply each function to its corresponding input value with \ensuremath{\Varid{listApp}} (Fig.~\ref{fig:listApp}) and then fork the computation away with \ensuremath{\Varid{arr}\;\Varid{spawnP}} inside a \ensuremath{\Varid{zipWithArr}} call. This yields a list \ensuremath{[\mskip1.5mu \Conid{Par}\;(\Conid{IVar}\;\Varid{b})\mskip1.5mu]}, which we then convert into \ensuremath{\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{sequenceA}}. In order to wait for the computation to finish, we map over the \ensuremath{\Conid{IVar}}s inside the \ensuremath{\Conid{Par}} monad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})}. The result of this operation is a \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} from which we can finally remove the monad again by running \ensuremath{\Varid{arr}\;\Varid{runPar}} to get our output of \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
\\
\>[9]{}\hsindent{8}{}\<[17]%
\>[17]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{zipWithArr}\;(\Varid{app}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{spawnP})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{arr}\;\Varid{sequenceA}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{arr}\;\Varid{runPar}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$ %% formatting
\caption{ArrowParallel instance for the Par Monad backend}
\label{fig:ArrowParallelParMonad}
\end{figure}

\subsubsection{Eden}
For both the Multicore Haskell and \ensuremath{\Conid{Par}} Monad implementations we could use general instances of \ensuremath{\Conid{ArrowParallel}} that just require the \ensuremath{\Conid{ArrowApply}} and \ensuremath{\Conid{ArrowChoice}} typeclasses. With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows. While we could still manage to have only one class in the module by introducing a typeclass: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Conid{ArrowUnwrap}\;\Varid{arr}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to (\Varid{a}\to \Varid{b}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{possible ArrowUnwrap typeclass}
% \label{fig:ArrowUnwrap}
% \end{figure}
We don't do it here for aesthetic resons, though. For now, we just implement \ensuremath{\Conid{ArrowParallel}} for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow \Conid{ArrowParallel}\;(\to )\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\;\Varid{as}\mathrel{=}\Varid{spawnF}\;\Varid{fs}\;\Varid{as}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{ArrowParallel instance for functions in the Eden backend}
% \label{fig:ArrowParallelEdenFns}
% \end{figure}
and the Kleisli type: % (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{Monad}\;\Varid{m},\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b},\Conid{Trans}\;(\Varid{m}\;\Varid{b}))\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;(\Conid{Kleisli}\;\Varid{m})\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\;\Varid{conf}\;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}(\Varid{arr}\mathbin{\$}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{map}\;(\lambda (\Conid{Kleisli}\;\Varid{f})\to \Varid{f})\;\Varid{fs}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}(\Conid{Kleisli}\mathbin{\$}\Varid{sequence}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
% \label{fig:ArrowParallelEdenKleisli}
% \end{figure}
